---
title: "thesis"
author: "Bin Yu"
date: "2025-02-01"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the data and cleaning

```{r}
rm(list = ls())

```

```{r}
library(frequency)
library(descr)
library(ggplot2)
load("/Users/yubin/Desktop/Thesis/WTC_COVID-selected/data/merged_main_2024-12-21.RData")
```

```{r}
d <- dd
rm(dd)
freq(d$Categorization)
## cleaning / recoding
# 99 (NA): missing or unable to categorize
# 0: asymptimatic
# 1: mild
# 2: moderate
# 3: severe
d$severity = 99
d$severity[d$Categorization=='Asymptomatic']=0
d$severity[d$Categorization=='asymptomatic']=0
d$severity[d$Categorization=='mild']=1
d$severity[d$Categorization=='Mild']=1
d$severity[d$Categorization=='moderate']=2
d$severity[d$Categorization=='Moderate']=2
d$severity[d$Categorization=='severe']=3
d$severity[d$Categorization=='Severe']=3
d$severity[d$Categorization=='SEVERE']=3
d$severity[d$severity==99] = NA
# new variable for severity
freq(d$severity)

```

```{r}
# 2 severity categories: asymptomatic/mild vs moderate/severe
## this is the main outcome variable
d$severity2mid=NA
d$severity2mid[d$severity<2]=0
d$severity2mid[d$severity>=2]=1
freq(d$severity2mid)
d$severity2mid = as.factor(d$severity2mid)
```

```{r}
## Long COVID 
freq(d$Any.residual.lingering.symptoms.due.to.previous.infection.)

# detailed variable
d$pacs = 99 # unknown
d$pacs[d$Any.residual.lingering.symptoms.due.to.previous.infection.=='No']=0

d$pacs[d$Any.residual.lingering.symptoms.due.to.previous.infection.=='Less than 4 weeks']=1

d$pacs[d$Any.residual.lingering.symptoms.due.to.previous.infection.=='1-2 months']=2
d$pacs[d$Any.residual.lingering.symptoms.due.to.previous.infection.=='1-2 Months']=2

d$pacs[d$Any.residual.lingering.symptoms.due.to.previous.infection.=='2-6 Months']=3
d$pacs[d$Any.residual.lingering.symptoms.due.to.previous.infection.=='2-6 months']=3
d$pacs[d$Any.residual.lingering.symptoms.due.to.previous.infection.=='2- 6 months']=3

d$pacs[d$Any.residual.lingering.symptoms.due.to.previous.infection.=='6-12 months']=4
d$pacs[d$Any.residual.lingering.symptoms.due.to.previous.infection.=='6-12 Months']=4

d$pacs[d$Any.residual.lingering.symptoms.due.to.previous.infection.=='Greater than 1 year']=5

d$pacs[d$Any.residual.lingering.symptoms.due.to.previous.infection.=='Greater than 2 years']=6

d$pacs[d$Any.residual.lingering.symptoms.due.to.previous.infection.=='Greater than 3 years']=7
d$pacs[d$Any.residual.lingering.symptoms.due.to.previous.infection.=='Greater than 4 years']=7

d$pacs[d$Any.residual.lingering.symptoms.due.to.previous.infection.=='Yes']=8

freq(d$pacs)

d$pacs = as.factor(d$pacs)
```

```{r}
# [THIS] binary: long COVID is >2 months, "yes" is "yes" (current definition is 3 months)
d$pacs_bin2m = 1
d$pacs_bin2m[d$pacs==0]=0
d$pacs_bin2m[d$pacs==1]=0
d$pacs_bin2m[d$pacs==2]=0
d$pacs_bin2m[d$pacs==99]=NA
```

```{r}
freq(d$pacs_bin2m)
```

```{r}

#####                           WTC exposures                              #####
# *****************************************************************************#


### Aggregate exposure (multi-level) 
# **************************************

### 4 categories of exposure: from lowest to highest
freq(d$exp4)
freq(d$exp4_imputed)
freq(d$Exp4_Imputed)# use imputed
#  cleaning / recoding" 0 = low; 1 = intermediate; 2 = high; 3 = very high
d$exp4_clean = d$exp4_imputed
freq(d$exp4_clean)
# make factor variables be factors
d$exp4_clean = as.factor(d$exp4_clean)



# exp4_binary: merge 0/1 and 2/3
d$exp4_binary = NA
d$exp4_binary[d$exp4_clean==0]=0
d$exp4_binary[d$exp4_clean==1]=0
d$exp4_binary[d$exp4_clean==2]=1
d$exp4_binary[d$exp4_clean==3]=1
freq(d$exp4_binary)




# exp4_3level: merge 2/3 (high and very high)
## this is the main variable in the analysis 
d$exp4_3level = NA
d$exp4_3level[d$exp4_clean==0]=0
d$exp4_3level[d$exp4_clean==1]=1
d$exp4_3level[d$exp4_clean==2]=2
d$exp4_3level[d$exp4_clean==3]=2
freq(d$exp4_3level)
# make factor variables be factors
d$exp4_3level = as.factor(d$exp4_3level)



# exp4_exclude: exclude 3 (very high)
d$exp4_exclude = d$exp4_clean
d$exp4_exclude[d$exp4_clean==3]=NA
freq(d$exp4_exclude)
d$exp4_exclude = as.factor(d$exp4_exclude)



### Exp5: N=74 re-classified from 'missing' to '1.5' level: doesn't add much
freq(d$Exp5_Imputed)

```

```{r}

### Arrival time
# **************************************

### timing of arrival to sites: category: 1 = Sep 11, 2 = Sep 12-13, 3 = Sep 14-30, 4 = Oct 1 and beyond ##
freq(d$arrival)
d$wtc.arrive = d$arrival-1
d$wtc.arrive = as.factor(d$wtc.arrive)
freq(d$wtc.arrive)

# 3 levels: merge Sep 12-13 and Sept 14-30 in one category
d$wtc.arrive3 = NA
d$wtc.arrive3[d$arrival==1] = 0
d$wtc.arrive3[d$arrival==2] = 1
d$wtc.arrive3[d$arrival==3] = 1
d$wtc.arrive3[d$arrival==4] = 2
d$wtc.arrive3 = as.factor(d$wtc.arrive3)
freq(d$wtc.arrive3)

# 2 levels: Sep 11 and after Sep 11
d$wtc.arrive2 = NA
d$wtc.arrive2[d$arrival==1] = 0
d$wtc.arrive2[d$arrival==2] = 1
d$wtc.arrive2[d$arrival==3] = 1
d$wtc.arrive2[d$arrival==4] = 1
freq(d$wtc.arrive2)
d$wtc.arrive2=as.factor(d$wtc.arrive2)


# arrival and dust: aggregate
# category: 1 = Sep 11 in the dust cloud, 2 = Sep 11 not in the dust cloud, 
# 3 = Sep 12-13, 4 = Sep 14-30, 5 = Oct 1 and beyond
freq(d$ArrivalDust)
# freq(d$ArrivalDust2020): adds category 1.5, not needed

d$wtc.arrive.dust3 = NA
d$wtc.arrive.dust3[d$ArrivalDust==1] = 0
d$wtc.arrive.dust3[d$ArrivalDust==2] = 1
d$wtc.arrive.dust3[d$ArrivalDust==3] = 2
d$wtc.arrive.dust3[d$ArrivalDust==4] = 2
d$wtc.arrive.dust3[d$ArrivalDust==5] = 3
d$wtc.arrive.dust3 = as.factor(d$wtc.arrive.dust3)
freq(d$wtc.arrive.dust3)
```

```{r}
### DUST
# **************************************

# First exposure to the WTC site or rescue/recovery effort whether or not to work/volunteer: 
# category: 1 = Sep 11 in the dust cloud, 1.5 = Sep 11 unknown dust cloud, 
# 2 = Sep 11 not in the dust cloud, 3 = Sep 12-13, 4 = Sep 14-30, 5 = Oct 1 and beyond
freq(d$dustfirstexposure)
d$dustfirstexposure.f = d$dustfirstexposure-1
d$dustfirstexposure.f = as.factor(d$dustfirstexposure.f)
freq(d$dustfirstexposure.f)



# caught-up in the DUST cloud from the collapse of the World Trade Center buildings
freq(d$dustc)
d$dust = d$dustc
d$dust = as.factor(d$dust)
d$dust[d$EAQ_08_04==0]=1
freq(d$dust)


# Worst exposure to dust: 
## 0 = Directly in the cloud of dust from the collapse of the World Trade Center buildings
## 1 = Exposed to significant amounts of dust but not directly in the cloud of dust from the collapse of the World Trade Center buildings
## 2 = Exposed to some dust but not in the cloud of dust from the collapse of the World Trade Center buildings
## 3 = Not exposed to dust and not in the cloud of dust from the collapse of the World Trade Center buildings
## 4 = Don't know
table(d$EAQ_08_04, useNA = 'ifany')
table(d$EAQ_08_04, d$dustc, useNA='ifany')


# Alternative measure of dust cloud exposure: binary
## 1 = In the dust cloud from buildings collapse OR exposed to significant amounts of dust, but not from buildings collapse
## 0 = Exposed to some dust or not exposed to dust
d$dust_alt = d$dust
d$dust_alt[d$EAQ_08_04==0]=1
d$dust_alt[d$EAQ_08_04==1]=1
freq(d$dust_alt)

table(d$dustc, d$dust_alt, useNA='ifany')


# Alternative measure of dust cloud exposure: 4 categories (doesn't improve compared to dust_alt)
d$dust_alt4 = 0 # 0 = not exposed to dust
d$dust_alt4[d$dust==1]=1
d$dust_alt4[d$EAQ_08_04==0]=1 # Directly in the cloud of dust from the collapse of the World Trade Center buildings
d$dust_alt4[d$EAQ_08_04==1]=2 # Exposed to significant amounts of dust but not directly in the cloud of dust from the collapse of the World Trade Center buildings
d$dust_alt4[d$EAQ_08_04==2]=3 # Exposed to some dust but not in the cloud of dust from the collapse of the World Trade Center buildings

d$dust_alt4 = as.factor(d$dust_alt4)
freq(d$dust_alt4)

```

```{r}

### PILE
# **************************************
# ever pile
freq(d$everpile)
d$everpile = as.factor(d$everpile)

# pile September
freq(d$pilesept)
d$pilesept = as.factor(d$pilesept)

```

```{r}


### DURATION of work in the response efforts
# **************************************
freq(d$wks11)
descr(d$wks11)
hist(d$wks11, breaks=30)
# categorize wks11
d$wks11_cat = NA
d$wks11_cat[d$wks11==0]=0
d$wks11_cat[d$wks11>0 & d$wks11<5]=1
d$wks11_cat[d$wks11>=5 & d$wks11<11]=2
d$wks11_cat[d$wks11>=11]=3
d$wks11_cat = as.factor(d$wks11_cat)
freq(d$wks11_cat)

```

```{r}

## INJURY
# **************************************
freq(d$any_wtc_injur)
d$injur = 0
d$injur[d$any_wtc_injur==1]=1
freq(d$injur)
d$injur = as.factor(d$injur)

# injuries to specific organs: skip, too sparse
freq(d$wtc_injur_head)
freq(d$wtc_injur_internal)
freq(d$wtc_injur_25_50)
freq(d$wtc_injur_mt50)



### traditional (0) (fire, police, emergency) vs 
# non-traditional (1) responders (construction workers, news reporters, doctors, etc)
# **************************************
freq(d$ntrad)
d$ntrad = as.factor(d$ntrad)


```

```{r}


#####                              Demographics                            #####
# *****************************************************************************#


### Education
freq(d$Education)

d$edu = 99 
d$edu[d$Education=="Grad/Professional School"] = 0
d$edu[d$Education=="BA/BS Degree"] = 0
d$edu[d$Education=="Associate Degree"] = 0

d$edu[d$Education=="Some College"] = 1
d$edu[d$Education=="Professional School"] = 1 
d$edu[d$Education=="Technical School"] = 1         

d$edu[d$Education=="High School Grad"] = 2

d$edu[d$Education=="Some High School"] = 3     
d$edu[d$Education=="Elem School"]=3

d$edu = as.factor(d$edu)
freq(d$edu)


### Employment: most recent before Mar 1 2020
# Many of these records are very old (10 years before 2020)
freq(d$WorkStatus.x)

d$empl=99
d$empl[d$WorkStatus.x=='1 = Working: Full-Time']=0
d$empl[d$WorkStatus.x=='105 = Sporadic/Seasonal/As Needed']=0
d$empl[d$WorkStatus.x=='106 = Volunteer']=0
d$empl[d$WorkStatus.x=='2 = Working: Part-Time']=0
d$empl[d$WorkStatus.x=='3 = Working: Unknown Amount']=0


d$empl[d$WorkStatus.x=='14 = Disabled/Medical Leave: Unknown WTC Relationship, Unknown Work Relationship']=1
d$empl[d$WorkStatus.x=='15 = Disabled/Medical Leave: Unknown WTC Relationship, Work-Related']=1
d$empl[d$WorkStatus.x=='4 = Disabled/Medical Leave: WTC-Related, Any Work Relationship']=1
d$empl[d$WorkStatus.x=='5 = Disabled/Medical Leave: Non-WTC-Related, Unknown Work Relationship']=1
d$empl[d$WorkStatus.x=='6 = Disabled/Medical Leave: Non-WTC-Related, Work-Related']=1
d$empl[d$WorkStatus.x=='7 = Disabled/Medical Leave: Non-WTC-Related, Non-Work-Related']=1
d$empl[d$WorkStatus.x=='8 = Retired']=1
d$empl[d$WorkStatus.x=='9 = Unemployed']=1

d$empl = as.factor(d$empl)
freq(d$empl)

# depreesion and illness
### Recode MHSQ_14_23 and MHSQ_16_21 to depression and illness, respectively
d$depression <- ifelse(is.na(d$MHSQ_14_23), 99, d$MHSQ_14_23)
d$illness <- ifelse(is.na(d$MHSQ_16_21), 99, d$MHSQ_16_21)
d$depression <- ifelse(d$depression %in% c(1,2,3), 1, d$depression)
# Check frequencies for the recoded variables
freq(d$depression)
freq(d$illness)
```

```{r}
### Recode Race to a binary variable: 0 for White, 1 for Nonwhite
d$race_binary <- ifelse(d$Race == "White", 0, 1)
# Check frequencies for the new binary race variable
freq(d$race_binary)
```


```{r}
d$OAD <- ifelse(!is.na(d$cert_OAD) & d$cert_OAD == "OAD", 1, 0)
```

```{r}
library(tableone)
vars <- c("OAD", "severity2mid", "pacs_bin2m", "age911", "edu", "empl",
          "Race", "Sex", "dust_alt", "wtc.arrive.dust3", "wks11_cat")

factorVars <- c("edu", "empl", "Race", "Sex", "dust_alt", "wtc.arrive.dust3", "wks11_cat", 
                "OAD", "severity2mid", "pacs_bin2m")

table1 <- CreateTableOne(vars = vars, 
                         strata = "exp4_binary", 
                         data = d, 
                         factorVars = factorVars)

print(table1, showAllLevels = TRUE)

```

```{r}

library(naniar)

d$empl <- addNA(d$empl)
d$edu <- addNA(d$edu)

empl_dummies <- model.matrix(~ empl - 1, data = d)
edu_dummies <- model.matrix(~ edu - 1, data = d)

colnames(empl_dummies) <- paste0("empl_", levels(d$empl))
colnames(edu_dummies)  <- paste0("edu_", levels(d$edu))

d <- cbind(d, empl_dummies, edu_dummies)
# Convert depression and illness to factors if they are not already
d$depression <- as.factor(d$depression)
d$illness    <- as.factor(d$illness)

# Create dummy variables for depression and illness
depression_dummies <- model.matrix(~ depression - 1, data = d)
illness_dummies    <- model.matrix(~ illness - 1, data = d)

# Rename columns for clarity
colnames(depression_dummies) <- paste0("depression_", levels(d$depression))
colnames(illness_dummies)    <- paste0("illness_", levels(d$illness))

# Add the new dummy variables to the data frame
d <- cbind(d, depression_dummies, illness_dummies)
```


After cleaning the data, show the table 1 again:

```{r}
library(tableone)
d$BMI <- as.numeric(as.character(d$BMI))

vars <- c("OAD", "severity2mid", "pacs_bin2m", "age911", "edu", "empl",
          "Race", "ntrad","depression","illness", "Sex", "BMI")

factorVars <- c("edu", "empl", "Race", "Sex", "OAD", "severity2mid", "pacs_bin2m","ntrad","depression","illness")

d <- d[complete.cases(d[, c("exp4_binary", "OAD", "severity2mid", "pacs_bin2m", "age911", "edu", "empl", "Race", "Sex", "dust_alt", "wtc.arrive.dust3", "wks11_cat","ntrad", "BMI")]), ]

table1 <- CreateTableOne(vars = vars, 
                         strata = "exp4_binary", 
                         data = d, 
                         factorVars = factorVars)

print(table1, showAllLevels = TRUE)
```

Traditional Model:

```{r}
library(lavaan)
# Models 
#   - exp4_binary as exposure
#   - OAD as the first mediator
#   - severity2mid as the second mediator
#   - pacs_bin2m as the results
#   - age911、edu、empl as the confounders
model <- '
  # mediation model 1: OAD predicted by exposure and confounders 
  OAD ~ a1*exp4_binary + age911 + edu_1 + edu_2 + edu_3 + edu_99 + empl_1 + empl_99

  # mediation model 2: severity predicted by exposure, confounders and OAD (mediator 1)
  severity2mid ~ a2*exp4_binary + b1*OAD + age911  + edu_1 + edu_2 + edu_3 + edu_99 + empl_1 + empl_99

  # results: long covid predicted by exposure, mediators and confounders
  pacs_bin2m ~ c_prime*exp4_binary + b2*OAD + b3*severity2mid + age911 + edu_1 + edu_2 + edu_3 + edu_99+ empl_1 + empl_99

  # define the mediation effects
  # chain effect：exp4_binary -> OAD -> severity2mid -> pacs_bin2m
  ind_chain := a1 * b1 * b3

  # indirect effect for exp4_binary to severity2mid 
  ind_med2  := a2 * b3

  # indirect effect of exp4_binary to OAD 
  ind_med1  := a1 * b2

  # Total indirect effect
  total_ind := ind_chain + ind_med2 + ind_med1

  # Total effect
  total_effect := c_prime + total_ind
'

fit <- sem(model,
           data = d,
           ordered = c("exp4_binary", "OAD", "severity2mid", "pacs_bin2m"),
           estimator = "WLSMV")

summary(fit, standardized = TRUE, fit.measures = TRUE)
```

## Overview

The results were obtained using the `lavaan` package with the DWLS estimator and NLMINB optimization. The model was fitted on 2228 observations after dropping NA in the exposure and outcomes. Overall, the model fit indices indicate a good fit (CFI = 1.000, TLI = 1.000, RMSEA = 0.000, SRMR = 0.000).

## Model Specification

The model is structured in three parts:

1.  **Mediator Model for OAD:**\
    $$
    \text{logit}\,\Pr\{ \text{OAD} = 1 \mid X, C \} = a_0 + a_1\, X + \gamma_1^\top C,
    $$ where $X$ (denoted as `exp4_bn`) is the exposure and $C$ represents covariates (age911, edu, empl, etc.). The estimate $a_1 = 0.286$ (p \< 0.001) indicates that a one-unit increase in $X$ increases the log-odds of OAD by 0.286.

2.  **Mediator Model for Severity2mid:**\
    $$
    \text{logit}\,\Pr\{ \text{severity2mid} = 1 \mid X, \text{OAD}, C \} = b_0 + a_2\, X + b_1\, \text{OAD} + \gamma_2^\top C,
    $$ where the effect of OAD on severity2mid is significant ($b_1 = 0.303$, p \< 0.001), while the direct effect of $X$ ($a_2 = 0.052$) is not significant.

3.  **Outcome Model for pacs_bin2m:**\
    $$
    \text{logit}\,\Pr\{ Y = 1 \mid X, \text{OAD}, \text{severity2mid}, C \} = c_0 + c' X + b_2\, \text{OAD} + b_3\, \text{severity2mid} + \gamma_3^\top C.
    $$ Here, the effect of severity2mid is strong and significant ($b_3 = 0.435$, p \< 0.001), while the direct effect of $X$ (denoted $c'$, estimated as 0.047) is not significant.

## Effect Decomposition

Defined parameters allow us to decompose the total effect of $X$ on $Y$ into indirect and direct components. In our chain mediation model, the chain indirect effect is given by the product: $$
\text{ind\_chain} = a_1 \times b_1 \times b_3,
$$ which is estimated as 0.038 (p \< 0.001).

Additionally, we have:

\- **ind_med1 (Indirect effect through OAD alone):** 0.017 (p = 0.138),

\- **ind_med2 (Indirect effect through severity2mid alone):** 0.023 (p = 0.449),

\- **Total Indirect Effect:** 0.077 (p = 0.017),

\- **Total Effect:** 0.124 (p = 0.082).

## Interpretation

-   **Chain Indirect Effect:**\
    The value of 0.038 indicates that there is a significant sequential (chain) mediation effect. This represents the effect of $X$ on $Y$ transmitted via OAD and then severity2mid: $$
    \text{Chain Indirect Effect} = a_1 \times b_1 \times b_3 \approx 0.286 \times 0.303 \times 0.435 \approx 0.038.
    $$

-   **Natural Direct Effect (NDE):**\
    The direct effect of $X$ on $Y$ (not transmitted through the mediators) is small and not significant ($c' \approx 0.047$, p = 0.502).

-   **Total Effect:**\
    The overall effect of $X$ on $Y$ is the sum of the direct and indirect effects, estimated as 0.124 (marginally non-significant, p = 0.082). The indirect pathway (combined mediation) plays a substantial role in transmitting the effect of $X$ on $Y$.

## Conclusion

The results support a chain mediation process in which the exposure $X$ influences the outcome $Y$ through its effect on OAD and subsequently on severity2mid. Although the natural direct effect is small, the significant chain indirect effect suggests that the sequential mediation pathway is an important mechanism in this model.

---

## Chain Mediation Analysis using the paths Package

In this analysis, we assume the following causal chain:
  
\[
\text{exp4\_binary} \longrightarrow \text{OAD} \longrightarrow \text{severity2mid} \longrightarrow \text{pacs\_bin2m}
\]

where  
- **exp4_binary** is the exposure,  
- **OAD** is the first mediator,  
- **severity2mid** is the second mediator, and  
- **pacs_bin2m** is the outcome.  

Covariates include **age911**, **edu_1**, **edu_2**, **edu_3**, **edu_99**, **empl_1**, and **empl_99**.

We will use GLMs to estimate the following models:

1. **Mediator 1 Model (OAD):**
\[
\text{logit}\,\Pr\{\text{OAD}=1 \mid \text{exp4\_binary}, C\} = \alpha_0 + \alpha_1\,\text{exp4\_binary} + \gamma_1^\top C,
\]
where \( C \) denotes the covariates.

2. **Mediator 2 Model (severity2mid):**
\[
\text{logit}\,\Pr\{\text{severity2mid}=1 \mid \text{exp4\_binary}, \text{OAD}, C\} = \beta_0 + \beta_1\,\text{exp4\_binary} + \beta_2\,\text{OAD} + \gamma_2^\top C.
\]

3. **Outcome Model (pacs\_bin2m):**
\[
\text{logit}\,\Pr\{Y=1 \mid \text{exp4\_binary}, \text{OAD}, \text{severity2mid}, C\} = \delta_0 + \delta_1\,\text{exp4\_binary} + \delta_2\,\text{OAD} + \delta_3\,\text{severity2mid} + \gamma_3^\top C.
\]

Additionally, we fit a propensity score model for the exposure:
\[
\text{logit}\,\Pr\{\text{exp4\_binary}=1 \mid C\} = \eta_0 + \eta^\top C.
\]

The following R code implements these steps using the **paths** package.

```{r, message=FALSE, warning=FALSE}
library(paths)

# Define the mediators as a list of variable names (chain mediation with two mediators)
mediators <- list("OAD", "severity2mid")
# Define formulas for each GLM
formula_m0 <- pacs_bin2m ~ exp4_binary + age911 + edu_1 + edu_2 + edu_3 + edu_99 + empl_1 + empl_99 + ntrad + depression_1 + depression_99 + illness_1 + illness_99 + BMI
# Mediator 1 model: OAD
formula_m1 <- pacs_bin2m ~ exp4_binary + age911 + edu_1 + edu_2 + edu_3 + edu_99 + empl_1 + empl_99 + ntrad + depression_1 + depression_99 + illness_1 + illness_99 + BMI + OAD

# Mediator 2 model: severity2mid
formula_m2 <- pacs_bin2m ~ exp4_binary + age911 + edu_1 + edu_2 + edu_3 + edu_99 + empl_1 + empl_99 + ntrad + depression_1 + depression_99 + illness_1 + illness_99 + BMI + OAD + severity2mid
# Propensity score model for the exposure
formula_ps <- exp4_binary ~ age911 + edu_1 + edu_2 + edu_3 + edu_99 + empl_1 + empl_99 + ntrad + depression_1 + depression_99 + illness_1 + illness_99 + BMI

# Fit the GLMs
glm_m0 <- glm(formula_m0, family = binomial("logit"), data = d)
glm_m1 <- glm(formula_m1, family = binomial("logit"), data = d)
glm_m2 <- glm(formula_m2, family = binomial("logit"), data = d)

glm_ymodels <- list(glm_m0, glm_m1, glm_m2)

# Fit the propensity score model
glm_ps <- glm(formula_ps, family = binomial("logit"), data = d)

# Run the causal paths analysis using the paths package.
# The exposure is 'exp4_binary' and the outcome is 'pacs_bin2m'.
# We use 1000 bootstrap replicates for inference.
paths_glm <- paths(a = "exp4_binary", y = "pacs_bin2m", m = mediators,
                   glm_ymodels, ps_model = glm_ps, data = d, nboot = 1000)

# View a summary of the paths analysis results
summary(paths_glm)

# Optionally, plot sensitivity analyses if desired
plot(paths_glm)
```
```{r}
library(paths)

# Define the mediators as a list of variable names (chain mediation with two mediators)
mediators <- list("OAD")
# Define formulas for each GLM
formula_m0 <- severity2mid ~ exp4_binary + age911 + edu_1 + edu_2 + edu_3 + edu_99 + empl_1 + empl_99
# Mediator 1 model: OAD
formula_m1 <- severity2mid ~ exp4_binary + age911 + edu_1 + edu_2 + edu_3 + edu_99 + empl_1 + empl_99 + OAD

# Propensity score model for the exposure
formula_ps <- exp4_binary ~ age911 + edu_1 + edu_2 + edu_3 + edu_99 + empl_1 + empl_99

# Fit the GLMs
glm_m0 <- glm(formula_m0, family = binomial("logit"), data = d)
glm_m1 <- glm(formula_m1, family = binomial("logit"), data = d)

glm_ymodels <- list(glm_m0, glm_m1)

# Fit the propensity score model
glm_ps <- glm(formula_ps, family = binomial("logit"), data = d)
paths_glm <- paths(a = "exp4_binary", y = "severity2mid", m = mediators,
                   glm_ymodels, ps_model = glm_ps, data = d, nboot = 1000)

# View a summary of the paths analysis results
summary(paths_glm)

# Optionally, plot sensitivity analyses if desired
plot(paths_glm)
```
```{r}
# Define the mediators as a list of variable names (chain mediation with two mediators)
mediators <- list("severity2mid")
# Define formulas for each GLM
formula_m0 <- pacs_bin2m ~ exp4_binary + age911 + edu_1 + edu_2 + edu_3 + edu_99 + empl_1 + empl_99
# Mediator 1 model: OAD
formula_m1 <- pacs_bin2m ~ exp4_binary + age911 + edu_1 + edu_2 + edu_3 + edu_99 + empl_1 + empl_99 + severity2mid

# Propensity score model for the exposure
formula_ps <- exp4_binary ~ age911 + edu_1 + edu_2 + edu_3 + edu_99 + empl_1 + empl_99

# Fit the GLMs
glm_m0 <- glm(formula_m0, family = binomial("logit"), data = d)
glm_m1 <- glm(formula_m1, family = binomial("logit"), data = d)

glm_ymodels <- list(glm_m0, glm_m1)

# Fit the propensity score model
glm_ps <- glm(formula_ps, family = binomial("logit"), data = d)

# Run the causal paths analysis using the paths package.
# The exposure is 'exp4_binary' and the outcome is 'pacs_bin2m'.
# We use 1000 bootstrap replicates for inference.
paths_glm <- paths(a = "exp4_binary", y = "pacs_bin2m", m = mediators,
                   glm_ymodels, ps_model = glm_ps, data = d, nboot = 1000)

# View a summary of the paths analysis results
summary(paths_glm)

# Optionally, plot sensitivity analyses if desired
plot(paths_glm)
```

```{r}
library(medflex)
vars_needed <- c("pacs_bin2m", "exp4_binary", "OAD", "severity2mid", "age911", "edu", "empl")
d_clean <- d[complete.cases(d[, vars_needed]), ]
```

## Chain Mediation Model Using medflex

In this analysis, we examine a chain (sequential) mediation model where the exposure variable, **exp4_binary**, influences the outcome **pacs_bin2m** through two mediators in sequence: **OAD** (mediator 1) and **severity2mid** (mediator 2). In our theoretical model, the causal chain is assumed to be:

$$
X \longrightarrow M_1 \longrightarrow M_2 \longrightarrow Y,
$$

where\
- $X$ is the exposure (exp4_binary),\
- $M_1$ is the first mediator (OAD),\
- $M_2$ is the second mediator (severity2mid),\
- $Y$ is the outcome (pacs_bin2m),\
- and $C$ represents the covariates (age911, edu, empl).

### Data Expansion

We begin by expanding the dataset using the `neImpute` function from the **medflex** package. The imputation model includes the exposure, the two mediators, their interaction term, and the covariates. The parameter `nMed = 2` indicates that the two variables immediately following the exposure in the formula (i.e., **OAD** and **severity2mid**) are treated as mediators. The imputation model is given by:

$$
\text{logit}\,\Pr\{Y = 1 \mid X, M_1, M_2, C\} = \alpha_0 + \alpha_1 X + \alpha_2 M_1 + \alpha_3 M_2 + \alpha_4 (M_1 \times M_2) + \gamma^\top C.
$$

This model allows the interaction between $M_1$ and $M_2$

## Natural Effects Model for Chain Mediation

After data expansion, the **medflex** package generates two counterfactual versions of the exposure: - **exp4_binary0** (denoted as $x^*$): the exposure under the control condition, - **exp4_binary1** (denoted as $x$): the exposure under the treatment condition.

We then fit a natural effects model on the expanded dataset. In the **conditional model** (which adjusts for covariates), the model is specified as:

$$
\text{logit}\,\Pr\{Y(x, M(x^*))=1 \mid C\} = \beta_0 + \beta_1\, x^* + \beta_2\, x + \beta_3\, (x^* \times x) + \gamma^\top C,
$$

where:

\- $x^*$ is the counterfactual exposure (corresponding to **exp4_binary0**),

\- $x$ is the actual exposure (corresponding to **exp4_binary1**),

\- $C$ represents the covariates (e.g., **age911**, **edu**, **empl**),

\- $\beta_1$ captures the effect transmitted through the mediators (i.e., the joint or chain indirect effect),

\- $\beta_2$ represents the natural direct effect (NDE) of the exposure on the outcome (i.e., the effect not transmitted through the mediators),

\- $\beta_3$ accounts for any interaction between $x^*$ and $x$, and

\- $\gamma$ is a vector of coefficients for the covariates.

In this formulation, the **total effect (TE)** on the log-odds scale is approximately $\beta_1 + \beta_2$ (plus any contribution from the interaction term, if significant).

For the **marginal model** (without adjusting for covariates), the specification is:

$$
\text{logit}\,\Pr\{Y(x, M(x^*))=1\} = \beta_0 + \beta_1\, x^* + \beta_2\, x + \beta_3\, (x^* \times x).
$$

### Effect Decomposition

We further decompose the estimated effects using the `neEffdecomp` function.

This decomposition yields:

\- **Natural Direct Effect (NDE):** The effect of $X$ on $Y$ that is not mediated by the mediators $M_1$ and $M_2$.

\- **Natural Indirect Effect (NIE):** The joint (chain) indirect effect, representing the effect transmitted sequentially through both mediators.

\- **Total Effect (TE):** The overall effect of $X$ on $Y$, which is the sum of the natural direct and joint indirect effects.

------------------------------------------------------------------------

Below is the corresponding R code used to fit these models:

```{r}

# random seed 
set.seed(23456)

# Step 1: expand the data
# except for confounders, we also include two mediators and their interaction term:
impData <- neImpute(
  pacs_bin2m ~ exp4_binary + OAD + severity2mid + OAD * severity2mid + age911 + edu + empl,
  family = binomial("logit"),
  nMed = 2,
  data = d_clean
)


# Step 2a: Conditional Natural Effects

neMod_joint_cond <- neModel(
  pacs_bin2m ~ exp4_binary0 * exp4_binary1 + age911 + edu + empl,
  family = binomial("logit"),
  expData = impData)

# results
summary(neMod_joint_cond)

# odds ratio:
exp(cbind(coef(neMod_joint_cond), confint(neMod_joint_cond)))

# Step 2b: Marginal Natural Effects

neMod_joint_marg <- neModel(
  pacs_bin2m ~ exp4_binary0 * exp4_binary1,
  family = binomial("logit"),
  expData = impData)

# results
summary(neMod_joint_marg)
exp(cbind(coef(neMod_joint_marg), confint(neMod_joint_marg)))

# Step 3a: decomposition of effects
effdecomp_cond <- neEffdecomp(neMod_joint_cond)
summary(effdecomp_cond)
# odds ratio
exp(coef(effdecomp_cond))

# Step 3b: marginal effects
effdecomp_marg <- neEffdecomp(neMod_joint_marg)
summary(effdecomp_marg)
exp(coef(effdecomp_marg))
```

Since the X-M interaction effect is not significant, try normal model:

```{r}

# expand the data
impData <- neImpute(
  pacs_bin2m ~ exp4_binary + OAD + severity2mid + OAD*severity2mid + age911 + edu + empl,
  family = binomial("logit"),
  nMed = 2,
  data = d_clean
)

# fit the natural effect model
neMod_joint <- neModel(
  pacs_bin2m ~ exp4_binary0 + exp4_binary1 + age911 + edu + empl,
  family = binomial("logit"),
  expData = impData,
  se = "bootstrap",
  nboot = 1000  # 1000 times of bootstrap
)
summary(neMod_joint)

# Optionally, convert the estimates to the odds ratio scale:
exp(cbind(coef(neMod_joint), confint(neMod_joint)))
# Step 2b: Marginal Natural Effects

neMod_marg <- neModel(
  pacs_bin2m ~ exp4_binary0 + exp4_binary1,
  family = binomial("logit"),
  expData = impData)

# results
summary(neMod_marg)
exp(cbind(coef(neMod_marg), confint(neMod_marg)))

# Step 3a: decomposition of effects
effdecomp_cond <- neEffdecomp(neMod_joint)
summary(effdecomp_cond)
# odds ratio
exp(coef(effdecomp_cond))

# Step 3b: marginal effects
effdecomp_marg <- neEffdecomp(neMod_marg)
summary(effdecomp_marg)
exp(coef(effdecomp_marg))



```

Also try to use mediation package to do the analysis ( not viable)

```{r}
library(mediation)

d_clean$exp4_binary <- as.numeric(as.character(d_clean$exp4_binary))
d_clean$severity2mid <- as.numeric(as.character(d_clean$severity2mid))

fit_M1 <- glm(OAD ~ exp4_binary + age911 + edu + empl,
              data = d_clean,
              family = binomial("logit"))

fit_M2 <- glm(severity2mid ~ exp4_binary + OAD + age911 + edu + empl,
              data = d_clean,
              family = binomial("logit"))

# exp4_binary -> OAD
med_out_M1 <- mediate(fit_M1, fit_M2,
                      treat = "exp4_binary", mediator = "OAD",
                      sims = 1000)
summary(med_out_M1)

# 
fit_Y <- glm(pacs_bin2m ~ exp4_binary + OAD + severity2mid + age911 + edu + empl,
             data = d_clean,
             family = binomial("logit"))

med_out_M2 <- mediate(fit_M2, fit_Y,
                      treat = "exp4_binary", mediator = "severity2mid",
                      sims = 1000)
summary(med_out_M2)

chain_ACME <- med_out_M1$d0 * med_out_M2$d0
chain_ACME
```